---
title: "cliqueR workflow playground"
output:
  html_notebook:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE)
```

This notebook walks the **R-side lineage-tracing workflow** end to end on the
`clique` toolkit:

```
GenBank ──clique genbank-to-yaml──▶ read-structure YAML
FASTQ  ──clique align────────────▶ BAM with `ce` (called-edit) tags
BAM    ──read_longread_lineage───▶ per-read table
table  ──build_indel_matrix──────▶ cell × site character matrix
matrix ──indel_distance──────────▶ distance ──tree_nj──▶ lineage tree
```

Run it chunk by chunk in RStudio. The steps that call the Rust binary skip
themselves gracefully if the binary or the example data isn't present, and the
final synthetic example always runs so you can poke at the tree-building on its
own.

# 1. Setup

Adjust `repo_dir` if this notebook isn't sitting in
`r_pkg/cliqueR/inst/examples/` of the clique monorepo.

```{r paths}
repo_dir <- normalizePath("../../../..", mustWork = FALSE)  # repo root
rust_dir <- file.path(repo_dir, "rust_cmd")
pkg_dir  <- file.path(repo_dir, "r_pkg", "cliqueR")

# Example data shipped with the Rust tree (used in sections 3–5).
read_structure <- file.path(rust_dir, "2026_02_08_maryam_RNF2_none_left_right.yml")
read1_full     <- file.path(rust_dir, "pR_NLR_8_S3_R1_001.fastq.gz")

# A scratch directory for generated files.
work_dir <- file.path(tempdir(), "cliqueR-play")
dir.create(work_dir, showWarnings = FALSE)
c(repo_dir = repo_dir, rust_dir = rust_dir, work_dir = work_dir)
```

Load `cliqueR` (installed if available, otherwise straight from the source
tree via `pkgload`):

```{r load-pkg}
# Prefer the live source tree so you always get the current code; fall back to
# an installed cliqueR.
if (dir.exists(pkg_dir) && requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(pkg_dir, quiet = TRUE)
} else {
  library(cliqueR)
}
```

Point `cliqueR` at the `clique` binary. It searches `CLIQUE_BIN`, then `PATH`,
then `rust_cmd/target/release/clique`; here we also fall back to a debug build.

```{r binary}
if (!nzchar(Sys.getenv("CLIQUE_BIN"))) {
  cands <- file.path(rust_dir, "target", c("release", "debug"), "clique")
  hit <- cands[file.exists(cands)]
  if (length(hit)) set_clique_binary(hit[1])
}

# TRUE if we can actually run the CLI sections below.
run_cli <- tryCatch({ clique_binary(); TRUE }, error = function(e) FALSE) &&
  file.exists(read_structure) && file.exists(read1_full)

if (run_cli) {
  message("clique binary: ", clique_binary())
} else {
  message("CLI sections will be skipped. Build the binary with: ",
          "cd rust_cmd && cargo build --release")
}
```

# 2. GenBank → read-structure YAML

`clique genbank-to-yaml` turns an annotated GenBank file into the read-structure
YAML the pipeline needs. A feature is picked up when a tag (default
`lineage_target`) appears in its name; UMI/barcode parameters and CRISPR target
types are read from `/clique_*` qualifiers. Here we write a tiny GenBank record
and convert it.

```{r genbank}
demo_gb <- file.path(work_dir, "demo.gb")
demo_yaml <- file.path(work_dir, "demo.yaml")

writeLines(c(
  "LOCUS       demo_amplicon             90 bp    DNA     linear   SYN 07-JUL-2026",
  "FEATURES             Location/Qualifiers",
  "     misc_feature    1..16",
  "                     /label=\"lineage_target cell barcode\"",
  "                     /clique_sort_type=\"KnownTag\"",
  "                     /clique_max_distance=\"2\"",
  "     misc_feature    17..28",
  "                     /label=\"lineage_target umi\"",
  "                     /clique_sort_type=\"DegenerateTag\"",
  "     misc_feature    30..52",
  "                     /label=\"lineage_target guide1\"",
  "                     /clique_type=\"Cas9WT\"",
  "     misc_feature    60..82",
  "                     /label=\"lineage_target guide2 Cas9ABE\"",
  "ORIGIN",
  "        1 aaaaaaaaaa aaaaaacccc ccccccccgG ACGGCTATAC AAGGCATCGC GGTTTTTTTC",
  "       61 TCGTCAATAC ACCTTACGGA GGAAAAAAAA",
  "//"
), demo_gb)

if (run_cli) {
  clique_run("genbank-to-yaml", list(
    genbank        = demo_gb,
    output         = demo_yaml,
    reference_name = "demo"
  ), echo = FALSE)
  cat(readLines(demo_yaml), sep = "\n")
} else {
  message("skipped (no binary)")
}
```

The barcode and UMI spans come back as symbol runs (`0000…`, `1111…`) in the
reference sequence — that's how the Rust extractor locates them.

# 3. Align reads → BAM

We align a small subset of the RNF2 example reads (a base-editing amplicon with
three near-identical references: `unedited`, `left`, `right`). Subsetting keeps
the demo fast; drop `n_reads` or point `read1_full` at your own FASTQ to scale
up.

```{r subset-fastq}
subset_fastq <- function(in_gz, out_gz, n_reads = 3000) {
  ci <- gzfile(in_gz, "rt"); on.exit(close(ci), add = TRUE)
  co <- gzfile(out_gz, "wt"); on.exit(close(co), add = TRUE)
  writeLines(readLines(ci, n = n_reads * 4L), co)
  out_gz
}

sub_fq      <- file.path(work_dir, "sub_R1.fastq.gz")
aligned_bam <- file.path(work_dir, "aligned.bam")

if (run_cli) {
  subset_fastq(read1_full, sub_fq, n_reads = 3000)
  clique_run("align", list(
    read1           = sub_fq,
    read_structure  = read_structure,
    output_bam_file = aligned_bam,
    threads         = 2
  ), echo = FALSE)
  message("wrote ", aligned_bam, " (", file.size(aligned_bam), " bytes)")
} else {
  message("skipped (no binary/data)")
}
```

# 4. Read the BAM into R

`read_longread_lineage()` streams the BAM through `samtools` and pulls the aux
tags. The one we care about is `ce` — the called-edit string per read.

```{r read-bam}
have_bam <- run_cli && file.exists(aligned_bam) && nzchar(Sys.which("samtools"))
if (have_bam) {
  lineage <- read_longread_lineage(aligned_bam, tags = c("ce"))
  cat("reads:", nrow(lineage), "\n")
  print(head(lineage, 4))
  cat("\nCalled-edit (ce) distribution — these are the ABE base edits:\n")
  print(head(sort(table(lineage$ce), decreasing = TRUE), 8))
} else {
  message("skipped (need a BAM and samtools on PATH)")
}
```

Each `ce` value is `_`-joined per target site; each site is `NONE` or
`&`-joined events like `1S+108+G` (an A→G base edit at reference position 108).

# 5. Indel matrix → tree (real data)

```{r real-tree, fig.width = 7, fig.height = 5}
if (have_bam) {
  # These references carry a single target, so the tree here is illustrative
  # (one site). Take a handful of reads spanning distinct edit states so the
  # plot stays legible; each read is treated as its own "cell".
  demo_reads <- lineage[!duplicated(lineage$ce), ]
  demo_reads <- head(demo_reads, 12)

  im <- build_indel_matrix(demo_reads)            # cell x site integer matrix
  cat("matrix:", nrow(im$matrix), "cells x", ncol(im$matrix), "site(s)\n")
  print(im$state_map[[1]])                         # edit string -> integer state

  d  <- indel_distance(im$matrix)
  tr <- tree_nj(d)
  plot(tr, cex = 0.6, main = "RNF2 subset (single-site, illustrative)")
} else {
  message("skipped (no BAM)")
}
```

# 6. A richer, self-contained example

Real lineage trees come from **many cut sites across many cells**. This block
needs no binary — it hand-builds a per-read table (as if it came from section 4)
with 3 sites and a nested clade structure, including a cell with several reads to
show majority-vote consensus.

```{r synthetic-tree, fig.width = 7, fig.height = 5}
lineage_syn <- data.frame(
  read_name = c("A1a", "A1b", "A1c", "A2", "A3",
                "B1", "B2", "B3", "C1", "C2"),
  cell      = c("A1", "A1", "A1", "A2", "A3",
                "B1", "B2", "B3", "C1", "C2"),
  reference = "amplicon",
  ce = c(
    "8D+12_2I+30+AC_NONE",   # A1 read 1
    "8D+12_2I+30+AC_NONE",   # A1 read 2 (agrees)
    "8D+12_NONE_NONE",       # A1 read 3 (site2 dropout -> outvoted)
    "8D+12_2I+30+AC_NONE",   # A2
    "8D+12_NONE_NONE",       # A3
    "NONE_NONE_5D+50",       # B1
    "NONE_NONE_5D+50",       # B2
    "NONE_NONE_NONE",        # B3 (all reference)
    "3D+11_NONE_5D+50",      # C1
    "3D+11_NONE_NONE"        # C2
  ),
  stringsAsFactors = FALSE
)

im  <- build_indel_matrix(lineage_syn, cell_col = "cell",
                          sites = c("site1", "site2", "site3"))
im$matrix
lapply(im$state_map, identity)

d   <- indel_distance(im$matrix)
tr  <- tree_nj(d, method = "bionj")
plot(tr, cex = 0.9, main = "Synthetic 3-site lineage")
```

Cell `A1` had three reads; the majority call (`8D+12_2I+30+AC_NONE`) wins, so its
site-2 dropout read doesn't skew the matrix.

# 7. Where to go next

The `tr` object is a standard `ape::phylo` (Newick-serializable with
`ape::write.tree(tr)`), which is exactly what the downstream tools consume:

- **SCOUT** (`hrstuart/SCOUT`) — Ornstein–Uhlenbeck modelling of gene expression
  along the tree; feed it `write.tree(tr)` plus a per-cell expression CSV.
- **Cassiopeia** — swap `tree_nj()` for `tree_cassiopeia(im$matrix, ...)` once
  that wrapper is wired up (Python via reticulate).
- **PATH** — trait-heritability testing on the same tree.

To scale to real single-cell data, use a read structure with a cell-barcode UMI
(so reads carry an `e0`-style tag) and pass `cell_col = "e0"` to
`build_indel_matrix()`; reads then collapse to one row per cell.
```
