---
title: "cliqueR: end-to-end example on the pR_NLR_8 test data"
output:
  html_document:
    toc: true
    toc_depth: 2
date: "`r Sys.Date()`"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

This walkthrough runs `cliqueR` against the `pR_NLR_8` paired-end FASTQ pair
that ships alongside the Rust source tree (`rust_cmd/pR_NLR_8_S3_R*_001.fastq.gz`).
The implemented layers (binary discovery, low-level CLI) execute live; the
higher-level R wrappers (IO, trees, metrics) are still stubs as of v0.0.0.9000
and their chunks are marked `eval = FALSE`.

**To re-run this end-to-end against your own data:** edit the `data_dir` and
filename variables in the *Paths* chunk and flip the `eval = FALSE` knob on
the CLI chunks. Expect a few minutes of wall-clock time on a typical laptop.

# 1. Setup

```{r libs}
library(cliqueR)
```

## Paths

Adjust these for your environment. Defaults point at the files in the sibling
`rust_cmd/` directory of this monorepo.

```{r paths}
# Default points at the sibling rust_cmd/ in this monorepo (four levels up
# from inst/examples/example-pR_NLR_8.Rmd). Override `data_dir` for your data.
data_dir <- normalizePath("../../../../rust_cmd", mustWork = FALSE)

read1          <- file.path(data_dir, "pR_NLR_8_S3_R1_001.fastq.gz")
read2          <- file.path(data_dir, "pR_NLR_8_S3_R2_001.fastq.gz")
read_structure <- file.path(data_dir, "2026_02_08_maryam_RNF2_none_left_right.yml")

work_dir       <- tempfile("cliqueR-example-")
dir.create(work_dir)
aligned_bam    <- file.path(work_dir, "pR_NLR_8.aligned.bam")
consensus_bam  <- file.path(work_dir, "pR_NLR_8.consensus.bam")

stopifnot(
  file.exists(read1),
  file.exists(read2),
  file.exists(read_structure)
)
list(read1 = read1, read2 = read2, read_structure = read_structure,
     work_dir = work_dir)
```

# 2. Locate the Rust binary

`cliqueR` looks at `CLIQUE_BIN`, then `PATH`, then
`rust_cmd/target/release/clique` relative to the current working directory.

```{r binary, error = TRUE}
clique_binary()
```

If the call above errored, build the binary or override the path:

```{r binary-override, eval = FALSE}
# Build once: cd ../../../rust_cmd && cargo build --release
set_clique_binary(file.path(data_dir, "target/release/clique"))
clique_binary()
```

## Sanity check: version banner

```{r version, eval = FALSE}
clique_run("--version", list(), echo = FALSE)$stdout
```

# 3. Align FASTQ → BAM

This invokes `clique align` directly via `clique_run()` — the typed `align()`
wrapper is still a stub.

```{r align, eval = FALSE}
res_align <- clique_run("align", list(
  read1                      = read1,
  read2                      = read2,
  read_structure             = read_structure,
  output_bam_file            = aligned_bam,
  threads                    = 4L,
  aligner                    = "WFA",
  min_read_length            = 50L,
  max_reference_multiplier   = 2L
))
res_align$status   # 0 on success
file.size(aligned_bam)
```

Once `align()` lands, the equivalent typed call will be:

```{r align-wrapped, eval = FALSE}
align(
  read1          = read1,
  read2          = read2,
  read_structure = read_structure,
  output_bam     = aligned_bam,
  threads        = 4L
)
```

# 4. Collapse aligned reads to per-UMI consensus

```{r collapse, eval = FALSE}
res_collapse <- clique_run("collapse", list(
  input_bam_file  = aligned_bam,
  output_bam_file = consensus_bam,
  read_structure  = read_structure,
  threads         = 4L,
  max_deletion    = 0L
))
res_collapse$status
file.size(consensus_bam)
```

Typed equivalent (when `collapse()` is implemented):

```{r collapse-wrapped, eval = FALSE}
collapse(
  input_bam      = aligned_bam,
  read_structure = read_structure,
  output_bam     = consensus_bam,
  threads        = 4L
)
```

# 5. Pull consensus reads into R

`pR_NLR_8` is a paired-end short-read sample. Use `read_10x_lineage()` when
you have a CellRanger output directory, or `read_longread_lineage()` for
PacBio / Nanopore amplicon BAMs.

```{r io, eval = FALSE}
# Long-read style: read every tag straight from the BAM
lineage <- read_longread_lineage(
  bam_path       = consensus_bam,
  read_structure = read_structure,
  tags           = c("CB", "UB")
)
head(lineage)
```

# 6. Build an indel character matrix

```{r indel-matrix, eval = FALSE}
sites <- c("site1", "site2", "site3")  # adjust to the read-structure YAML
im <- build_indel_matrix(lineage, sites = sites, missing_threshold = 0.5)
dim(im$matrix)
head(im$state_map[[sites[1]]])
```

# 7. Reconstruct a tree

Three solver families, same input:

```{r trees, eval = FALSE}
d        <- indel_distance(im$matrix)
tree_nj_ <- tree_nj(d, method = "bionj")
tree_par <- tree_parsimony(im$matrix)
tree_cas <- tree_cassiopeia(im$matrix, solver = "greedy")
```

# 8. Visualize

```{r viz, eval = FALSE}
plot_lineage_tree(
  tree_nj_,
  metadata = cell_meta,   # tibble with cell_barcode column
  color_by = "cluster",
  layout   = "circular"
)
```

# 9. Compare against a ground-truth tree

For benchmarking with simulated data or known fate maps:

```{r metrics, eval = FALSE}
rf_distance(tree_nj_, ground_truth)
triplet_correctness(tree_nj_, ground_truth, n_triplets = 1e5)
quartet_distance(tree_nj_, ground_truth)
ancestor_recall(tree_nj_, ground_truth)
```

# 10. Inspect intermediate artifacts

The work directory persists until the R session ends — useful for debugging
a failed run:

```{r workdir, eval = FALSE}
list.files(work_dir, recursive = TRUE)
```

```{r cleanup, include = FALSE, eval = FALSE}
unlink(work_dir, recursive = TRUE)
```

# Session info

```{r session}
sessionInfo()
```
