---
title: "Clique BAM output and auxiliary tags"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clique BAM output and auxiliary tags}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Clique writes standard BAM records plus auxiliary tags that preserve extracted
captures, event calls, molecule depth, and consensus provenance. This vignette
documents the fields most useful from R. The Rust CLI README is the
authoritative format contract.

## Import records into R

```{r}
library(cliqueR)

reads <- read_longread_lineage(
  "sample.consensus.bam",
  tags = c("e0", "o0", "ce", "pe", "rc", "dc", "ar", "rm")
)
```

The function returns one row per BAM record:

- `read_name` is BAM `QNAME`;
- `reference` is BAM `RNAME`; and
- requested auxiliary tags become character columns.

Missing tags are `NA`. Values are not type-converted:

```{r}
reads$rc <- as.integer(reads$rc)
reads$dc <- as.integer(reads$dc)
reads$rm <- as.numeric(reads$rm)
```

Passing `tags = NULL` discovers auxiliary tags from the first record and always
adds `ce`. Pass explicit names when tags may vary by record.

## Capture tags

Each YAML `umi_configurations` entry has a single-digit `symbol`. The symbol,
not the human-readable configuration name, determines the BAM tag:

```yaml
umi_configurations:
  cell_barcode:
    symbol: '0'
    sort_type: KnownTag
    length: 16
    order: 0
    max_distance: 2
  molecule_umi:
    symbol: '1'
    sort_type: DegenerateTag
    length: 12
    order: 1
    max_distance: 2
```

This layout produces `e0`/`o0` for the cell barcode and `e1`/`o1` for the
molecule UMI.

| Tag | Stage | Meaning |
| --- | --- | --- |
| `e<symbol>` | Align | Value extracted from that input read |
| `e<symbol>` | Collapse | Corrected value used to define the molecule group |
| `o<symbol>` | Collapse | Value from the representative selected read before correction |

A `KnownTag` capture can be corrected to its allowlisted value. A
`DegenerateTag` capture can be corrected to its cluster representative.
Consequently, `e<symbol>` and `o<symbol>` need not match in a collapsed BAM.

## Consensus depth and provenance

| Tag | Meaning |
| --- | --- |
| `rc` | Number of reads assigned to the corrected molecule group |
| `dc` | Number selected to build the consensus after downsampling |
| `ar` | Comma-separated selected input QNAMEs, in selection order |

For a normally collapsed record:

```text
length(strsplit(ar, ",")[[1]]) == dc <= rc
```

For example, `rc = 10`, `dc = 4`, and four names in `ar` mean ten reads formed
the corrected group and four were used for consensus construction. The other
six were omitted by `--maximum-reads-before-downsampling`.

The consensus `QNAME` is the first selected read name, not a generated molecule
identifier. `ar` contains selected reads only; it is not a complete list of all
`rc` group members after downsampling.

In `--correct-only` mode, clique emits one corrected record per retained input
record. In that mode `rc`, `dc`, and the number of names in `ar` are all one.

Summarize molecule support in R:

```{r}
support <- transform(
  reads,
  rc = as.integer(rc),
  dc = as.integer(dc)
)

summary(support$rc)
sum(support$rc - support$dc, na.rm = TRUE)
```

## Alignment and event tags

| Tag | Meaning |
| --- | --- |
| `rm` | Fraction of called, non-`N`, non-capture reference bases matching the output alignment |
| `as`, `rs` | Clique internal alignment scores; these are not standard SAM `AS` scores |
| `ce` | Chemistry-filtered CRISPR event calls |
| `pe` | Prime-edit classification when a reference has `PrimeEdit` targets |

The `ce` tag uses 0-based ungapped reference coordinates:

| Event | Encoding |
| --- | --- |
| Deletion | `<length>D+<reference-position>` |
| Insertion | `<length>I+<reference-position>+<inserted-bases>` |
| Substitution | `1S+<reference-position>+<alternate-base>` |
| No qualifying event | `NONE` |

Multiple events at one target are joined with `&`. Target fields are joined
with `_` in the same order as `targets` in the read-structure YAML. Event
filtering depends on the target chemistry, so `ce` is not an unfiltered list of
all alignment differences.

`build_indel_matrix()` splits `ce` on `_`. It intentionally keeps an
`&`-joined multi-event allele together as one character state.

For each `PrimeEdit` target, `pe` can contain:

- `WT`
- `PRECISE`
- `PARTIAL`
- `PRECISE_PLUS_BYPRODUCT`
- `SCAFFOLD_INCORPORATION`
- `INDEL`
- `OTHER`
- `NO_CALL`

Non-prime targets in a mixed target list are represented by `NA`. As with `ce`,
fields follow YAML target order.

## Reference-routing diagnostics

An alignment record can also include diagnostics from an optional reference
classifier:

| Tags | Classifier and meaning |
| --- | --- |
| `dm`, `di`, `da` | Discriminating-position margin, informative positions, ambiguity |
| `ib`, `im`, `ik`, `ia` | IDF best score, margin ratio, informative k-mers, ambiguity |
| `pb`, `pm`, `pi`, `pa` | POA best branch-vote score, margin, informative columns, ambiguity |

These describe reference assignment on align output and are not
collapse-group annotations.

## SAM conventions

- Alignment `QNAME` is the input read name; consensus `QNAME` is the first
  selected read name.
- Clique normalizes sequences into reference orientation. `FLAG` is currently
  `0`, so reverse complementation is not represented by SAM flag `0x10`.
- `MAPQ` is unset (`255`) and must not be interpreted as mapping confidence.
- `RNAME`, `POS`, and `CIGAR` refer to a configured amplicon, not necessarily a
  genomic chromosome.
- Clique output order is not guaranteed to be coordinate-sorted, and the CLI
  does not automatically create a BAM index.

Inspect raw records with:

```sh
samtools view -H sample.consensus.bam
samtools view sample.consensus.bam | cut -f1-6,12- | head
```

Sort and index when a downstream tool requires coordinate order:

```sh
samtools sort -o sample.consensus.sorted.bam sample.consensus.bam
samtools index sample.consensus.sorted.bam
```
