---
title: "Getting started with cliqueR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with cliqueR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`cliqueR` provides R-side configuration, execution, import, and analysis tools
for the `clique` Rust CLI. The CLI does the compute-intensive work: reference
assignment, alignment, capture extraction/correction, consensus generation,
and CRISPR event calling.

```{r}
library(cliqueR)
```

## Prerequisites

Build the CLI from the repository root:

```sh
cargo build --release --manifest-path rust_cmd/Cargo.toml
export CLIQUE_BIN="$PWD/rust_cmd/target/release/clique"
```

Install `samtools` separately. Pipeline scripts use it to sort and index BAMs,
and `read_longread_lineage()` uses `samtools view` to stream records into R.

Check which executable R will use:

```{r}
clique_binary()

# Override discovery for this R session, if needed.
set_clique_binary("/absolute/path/to/clique")
```

## Create a read-structure YAML

A read structure embeds each reference amplicon, target sequence, target
chemistry, read orientation, and capture/UMI layout. For a basic layout, create
it from FASTA records and a target table:

```{r}
targets <- data.frame(
  target = c(
    "GACGGCTATACAAGGCATCGCGG",
    "CTCGTCAATACACCTTACGGAGG"
  ),
  type = c("Cas9WT", "Cas9ABE")
)

matches <- fasta_targets_to_yaml(
  fasta_file = "amplicons.fa",
  targets = targets,
  output_file = "read_structure.yaml",
  merge = "ConcatenateBothForward",
  known_strand = TRUE,
  reads = list(
    list(type = "Read1", orientation = "Forward"),
    list(type = "Read2", orientation = "ReverseComplement")
  )
)
```

The helper searches both strands and writes only target strings that occur in
the forward reference. It reports which references contain each target.

The generated file is a starting point. Edit it directly when the experiment
needs:

- `umi_configurations` for known or degenerate captures;
- `target_locations` to select a repeated target occurrence;
- `prime_edits` specifications;
- index reads or spacers; or
- per-reference differences that cannot be inferred from the FASTA.

The full schema is documented in the Rust CLI README. Validate a finished
layout before a large run by processing a small read subset.

## Generate an align/collapse run

The current recommended high-level path is `generate_clique_script()`. It
writes a shell script so the complete command, input paths, and parameters can
be inspected and retained with an analysis.

```{r}
generate_clique_script(
  read_structure = "read_structure.yaml",
  reference = "amplicons.fa",
  read1 = "sample_R1.fastq.gz",
  read2 = "sample_R2.fastq.gz",
  output_script = "run_sample.sh",
  output_dir = "results",
  sample_name = "sample",
  threads = 8,
  collapse = TRUE
)
```

Run it from a shell:

```sh
bash run_sample.sh
```

For `sample_name = "sample"` and `collapse = TRUE`, the script produces:

| Output | Description |
| --- | --- |
| `results/sample.aligned.bam` | Direct `clique align` output |
| `results/sample.sorted.bam` | Coordinate-sorted alignment |
| `results/sample.sorted.bam.bai` | Index used by region-aware tools |
| `results/sample.consensus.bam` | One consensus per retained molecule group |

Set `correct_only = TRUE` to correct capture tags without generating
consensuses. Set `execute = TRUE` only when the R process should run the script
immediately.

The typed `align()` and `collapse()` functions are reserved APIs in this
development version and currently error. Use the script generator or
`clique_run()` instead.

## Use CLI flags not exposed by the generator

`clique_run()` accepts one subcommand and a named list:

```{r}
result <- clique_run("collapse", list(
  input_bam_file = "results/sample.sorted.bam",
  output_bam_file = "results/sample.consensus.bam",
  read_structure = "read_structure.yaml",
  threads = 8L,
  maximum_reads_before_downsampling = 100L,
  correct_only = FALSE
))

result$status
result$command
```

Underscores in names become dashes, so `input_bam_file` becomes
`--input-bam-file`. `TRUE` creates a bare flag; `FALSE`, `NULL`, and `NA` are
omitted. A nonzero process exit raises an R error containing the end of stderr.

## Read a clique BAM

Select the tags needed for the analysis explicitly:

```{r}
reads <- read_longread_lineage(
  "results/sample.consensus.bam",
  tags = c("e0", "e1", "ce", "pe", "rc", "dc", "ar")
)

head(reads)
```

The result contains `read_name`, `reference`, and one character column per tag.
Missing tags are `NA`. Convert numeric tags before summarizing:

```{r}
reads$rc <- as.integer(reads$rc)
reads$dc <- as.integer(reads$dc)

aggregate(rc ~ reference, reads, sum)
```

Use `region = "reference_name"` to import one reference from an indexed BAM.
When `tags = NULL`, the function discovers tags from the first record only;
explicit selection is safer for optional annotations such as `pe`.

See `vignette("bam-output", package = "cliqueR")` for the tag contract and
event encodings.

## Build a cell-by-target character matrix

The `ce` tag contains one event field per configured target. Group records by a
capture tag such as `e0`:

```{r}
characters <- build_indel_matrix(
  reads,
  sites = c("target_1", "target_2"),
  cell_col = "e0",
  event_col = "ce",
  missing_threshold = 0.5
)

characters$matrix
characters$state_map
```

At each target, `NONE` is state `0`, distinct edited alleles receive positive
integer states, and missing calls are `-1`. If a cell has multiple records,
the majority event string is selected independently at each target.

## Reconstruct a tree

For a fast exploratory tree, compute clique's missing-aware Hamming distance
and use neighbor-joining:

```{r}
distance <- indel_distance(characters$matrix)
tree <- tree_nj(distance, method = "bionj")
```

The package also supports unordered parsimony in R and four optional external
backends:

```{r}
# Unordered maximum parsimony with phangorn
parsimony_tree <- tree_parsimony(characters$matrix)

# IQ-TREE 2 maximum likelihood (MORPH/MK model)
iqtree <- tree_iqtree(
  characters$matrix,
  model = "MK+FQ",
  threads = 8,
  output_prefix = "results/sample.iqtree"
)

# VINE variational inference; request access to posterior samples
vine_tree <- tree_vine(
  characters$matrix,
  threads = 8,
  nsamples = 500,
  return_posterior = TRUE,
  output_prefix = "results/sample.vine"
)
vine_samples <- attr(vine_tree, "clique_posterior")

# Cassiopeia through reticulate
cassiopeia_tree <- tree_cassiopeia(
  characters$matrix,
  solver = "greedy"
)

# PHYLIP MIX with irreversible edit gains
mix_tree <- tree_mix(
  characters$matrix,
  method = "camin-sokal",
  output_prefix = "results/sample.mix"
)
```

| Function | Inference | Dependency and constraints |
| --- | --- | --- |
| `tree_nj()` | NJ/BIONJ on clique distances | R package `ape` |
| `tree_parsimony()` | Unordered parsimony ratchet | R packages `ape`, `phangorn` |
| `tree_iqtree()` | Maximum likelihood | IQ-TREE 2; maximum 32 observed states per site |
| `tree_vine()` | Variational Bayesian CRISPR model | VINE; ultrametric; mean tree plus optional posterior samples |
| `tree_cassiopeia()` | Greedy, ILP, hybrid, UPGMA, or NJ | Python Cassiopeia; ILP needs Gurobi |
| `tree_mix()` | Camin-Sokal or Wagner parsimony | PHYLIP MIX; edit states are one-hot expanded |

Every default result is an `ape::phylo` whose tips are the original cell names.
Set `output_prefix` for external command wrappers when the converted inputs,
logs, posterior samples, or native outputs should be retained. Without it,
temporary files are removed after the tree is imported.

## Plot a tree

Any of the trees above can be visualized through the same interface:

```{r}

cell_metadata <- data.frame(
  cell_barcode = rownames(characters$matrix),
  week = c("week0", "week1", "week2", "week3")
)

plot_lineage_tree(
  tree,
  annotations = cell_metadata,
  mapping_column = "cell_barcode",
  columns = "week",
  layout = "circular",
  tip_labels = FALSE
)
```

`tree_nj()` requires the optional CRAN package `ape`. Tree plotting requires
`ggplot2`, `ggtree`, and, for annotation rings, `ggnewscale`.

Tree-comparison metrics and 10x/CellRanger import remain documented reserved
APIs and are not implemented in this development release.
